A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-14 paper/paper.tex
Update the if rule rendering
Add the admissible pi alt rule in natural deduction form

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation. % In our presented proof, we use the informal
% language of sets, where the use of impredicativity feels more natural
% and less noticeable.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion}, a scheme in which logical
relation for dependent types is typically defined~\citep{todo}, we opt
for a direct explanation in informal language of sets, where
impredicativity manifests in the form of second-order logical
formulas and thus more intuitive to understand for readers with
general mathematics background.

@@

\begin{figure}[ht]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
now show that the same introduction rules for $[[Interp I i A S]]$ is
admissible for $[[InterpR i A S]]$. We give as example TODO:

In most informal presentations, instead of defining the logical
relation in two steps as we have shown above, the rules for $[[InterpR
i A S]]$ is given directly, with the implicit understanding of
inductive definitions nested inside a recursive function. We choose
the more explicit definition not only because it is directly definable
in existing proof assistants where inductive definitions must appear
at the top level, but also because it makes clear the induction
principle we are allowed to use when reasoning about $[[InterpR i A
S]]$.

The most general format of the induction principle over $[[InterpR i A
S]]$ is first by strong induction over the universe level $[[i]]$
followed by structural induction over $[[Interp I i A S]]$. As
examples, ... (\rref{I-Set} and \rref{I-Red}).

@@

\Rref{I-Pi} is the most interesting. The precondition consists of a
mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
subject to the constraint that $[[forall a, (# a in S implies Interp I
i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
the derivation, the statements from the precondition are equivalent to
the conjunction of the following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
\end{itemize}
In general, let $R$ be a binary relation over the sets $[[A]]$ and
$[[B]]$, it easy to verify that the following equivalence:
\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
  \in R\]

The left-hand side is equivalent to the right-hand side, but has no
mentioning of any functions. In the context of \rref{I-Pi}, through
the equivalence, we can more easily tell what's happening: $[[Pi A
B]]$ has a semantic interpretation of $[[A]]$ can be semantically
interpreted as the set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has a semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. If we used the alternative representation of the preconditions, we
would be forced to define our logical relation in the following way.

TODO: put the wrong logical relation that violates ...

Unfortunately, both versions not only violate the syntactic strict
positivity constraint required in proof assistants, but
are genuinely non-monotone over the domain of
relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
more intuitive introduction rules can be later shown to be admissible
a posteriori.

@@

Lemma~\ref{lemma:logreldeter} enables us to derive the following
alternative introduction and inversion principles for \rref{I-Pi}.
\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
  \label{lemma:piintroalt}
  Suppose the following two statements hold:
  \begin{itemize}
  \item $[[Interp I i A S]]$
  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
  \end{itemize}
  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
\end{lemma}
\begin{proof}
  Let $[[F]]$ be the relation defined as follows:
  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
  By the second bullet from the premise and
  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
  on the set $[[S]]$. The conclusion then trivially follows from
  \rref{I-Pi}.
\end{proof}

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
\label{sec:extension}
\begin{figure}[h]
  \[
    \begin{array}{lcl}
      % \beta\text{-}\mathit{neutral\ terms}\\
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      % \beta\text{-}\mathit{normal\ terms} \\
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & |   & [[\\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
\begin{figure}[h]
  \caption{Extended logical relation}
  \label{fig:logrelopen}
\end{figure}

>>>>

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation. % In our presented proof, we use the informal
% language of sets, where the use of impredicativity feels more natural
% and less noticeable.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion-dybjer}, a scheme in which
semantic models (including logical relations) for dependent types can
be defined, we opt for a direct explanation in informal language of
sets, where impredicativity manifests in the form of second-order
logical formulas and thus more intuitive to understand for readers
with general mathematics background.

@@

\begin{figure}[ht]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
now show that the same introduction rules for $[[Interp I i A S]]$ are
admissible for $[[InterpR i A S]]$, by
instantiating $[[I]]$ with $[[I i]] := [[{A | exists S , InterpR i A
  S}]]$, the same function $[[I]]$ used in the definition of $[[InterpR i A
S]]$. We give as example the following rules.
\begin{center}
\drule[]{IR-Void} \qquad \drule[]{IR-Set}
\end{center}

In most informal presentations, instead of defining the logical
relation in two steps as we have shown above, the rules for $[[InterpR
i A S]]$ are given directly, with the implicit understanding that the
relation is an inductive definition nested inside a recursive
function over the universe level $[[i]]$. We choose
the more explicit definition not only because it is directly definable
in existing proof assistants where inductive definitions must appear
at the top level, but also because it makes clear the induction
principle we are allowed to use when reasoning about $[[InterpR i A
S]]$.

% The paragraph below might be useful but I don't know how to phrase
% it well

% The most general format of the induction principle over
% $[[InterpR i A S]]$ is first by strong induction over the universe level $[[i]]$
% followed by structural induction over $[[Interp I i A S]]$. As
% examples, ... (\rref{I-Set} and \rref{I-Red}).

@@

\Rref{I-Pi} is the most interesting. The precondition consists of a
mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
subject to the constraint that $[[forall a, (# a in S implies Interp I
i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
the derivation, the statements from the precondition are equivalent to
the conjunction of the following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
\end{itemize}
In general, let $R$ be a binary relation over the sets $[[A]]$ and
$[[B]]$, it easy to verify the following equivalence, which is used to
justify the skolemization process~\citep{skolemization}.
\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
  \in R\]
The left-hand side is equivalent to the right-hand side, but has no
mentioning of any functions.
In the context of \rref{I-Pi}, through
the equivalence, we can more easily tell what's happening: the
function type $[[Pi A B]]$ has a semantic interpretation if its input
type $[[A]]$ can be semantically
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has a semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction
principle. However, as we will see in Lemma~\ref{lemma:piintroalt} to
show the admissibility of \rref{I-Pi} a posteriori.

@@

Lemma~\ref{lemma:logreldeter} enables us to show the admissibility of
\rref{I-PiAlt} and its related inversion principle.
\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
  \label{lemma:piintroalt}
  Suppose the following two statements hold:
  \begin{itemize}
  \item $[[Interp I i A S]]$
  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
  \end{itemize}
  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
\end{lemma}
\begin{proof}
  Let $[[F]]$ be the relation defined as follows:
  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
  By the second bullet from the premise and
  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
  on the set $[[S]]$. The conclusion then trivially follows from
  \rref{I-Pi}.
\end{proof}

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
\label{sec:extension}
\begin{figure}[h]
  \[
    \begin{array}{lcl}
      % \beta\text{-}\mathit{neutral\ terms}\\
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      % \beta\text{-}\mathit{normal\ terms} \\
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & |   & [[\\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
\begin{figure}[h]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation}
  \label{fig:logrelopen}
\end{figure}
